\documentclass[titlepage,numbers=noenddot,headinclude,
                footinclude=true,abstractoff,
                BCOR=5mm,paper=letter,fontsize=11pt,
                ngerman,american,dottedtoc]{scrreprt}

\input{linux_introduction-config}

\addbibresource{Bibliography.bib}
\addbibresource[label=ownpubs]{AMiede_Publications.bib}

%\hyphenation{put special hyphenation here}

\makeindex

\begin{document}
\frenchspacing
\raggedbottom
\selectlanguage{american} % american ngerman
%\renewcommand*{\bibname}{new name}
%\setbibpreamble{}
\pagenumbering{roman}
\pagestyle{plain}

%********************************************************************
% Frontmatter
%*******************************************************
%\include{FrontBackmatter/DirtyTitlepage}
\include{FrontBackmatter/Titlepage}
%\include{FrontBackmatter/Titleback}
%\cleardoublepage\include{FrontBackmatter/Dedication}
%\cleardoublepage\include{FrontBackmatter/Foreword}
%\cleardoublepage\include{FrontBackmatter/Abstract}
%\cleardoublepage\include{FrontBackmatter/Publications}
%\cleardoublepage\include{FrontBackmatter/Acknowledgments}
\pagestyle{scrheadings}
\cleardoublepage\include{FrontBackmatter/Contents}

%********************************************************************
% Mainmatter
%*******************************************************
\cleardoublepage
\pagenumbering{arabic}
%\setcounter{page}{90}
% use \cleardoublepage here to avoid problems with pdfbookmark
\cleardoublepage
\ctparttext{In the following chapters, we explain what an operating system is, as well as its relationship with the computer hardware and software applications. Following, we present a brief overview of Linux's history. Finally, we discuss what Linux distributions are, and we comment on some of the defining characteristics of some of the most widely used distributions.}
\part{Introduction to Command Line Interfaces}
\include{Chapters/Chaptera}
\include{Chapters/Chapterb}
\include{Chapters/Chapterc}
\cleardoublepage
\ctparttext{In the following chapters, we introduce the concept of command line interfaces, and explain how they can be accessed, as well as their relationship with the \textbf{shell}. Following, we present some basic \textbf{shell} commands together with methods to get help about these commmands. Finally, we end this part by introducing a series of tools that can be used to read and edit text files using a command line interface.}
\part{Introduction to Command Line Interfaces}
\include{Chapters/Chapter01}
\include{Chapters/Chapter02}
\include{Chapters/Chapter03}
\include{Chapters/Chapter04}
\include{Chapters/Chapter05}
\cleardoublepage
\ctparttext{In the following chapters, we discuss Linux file systems, covering both its formatting, as well as its directory hierarchy. Also, we introduce file links, discussing both hard links and soft links, as well as their properties. Following, we introduce file globbing techniques and discuss different wildcards that are frequently used. We also provide a gentle introduction to regular expressions, combined with the \mycommand{grep} tool. Finally, we discuss important filters and tools, such as \mycommand{cat}, \mycommand{sort}, \mycommand{gawk}, \mycommand{rename}, and \mycommand{find}. We finish this part by explaining the important concepts of piping and redirection.}
\part{Advanced Command Line Interfaces}
\include{Chapters/Chapter06}
\include{Chapters/Chapter07}
\include{Chapters/Chapter08}
\include{Chapters/Chapter09}
\include{Chapters/Chapter10}
\include{Chapters/Chapter11}
\include{Chapters/Chapter12}
\cleardoublepage
\ctparttext{In the following chapters, we introduce the concept of scripting, i.e., creating files containing a sequence of instructions that perform a specific task. Scripts can be used for automating repetitive tasks. They are widely used by system administrators to take care of tasks such as creating hundreds of new users, or configuring multiple machines. With scirpts, such tasks cab be done in a matter of minutes, instead of taking hours of days of repetitive work.}
\part{Scripting}
\include{Chapters/intro_scripting}
\include{Chapters/logical_expressions}
\include{Chapters/loops}
\include{Chapters/functions}
\include{Chapters/arrays}
\include{Chapters/permissions}
\include{Chapters/users_groups}
%\addtocontents{toc}{\protect\clearpage} % <--- just debug stuff, ignore
%\include{Chapters/Chapter03}
%\include{Chapters/Chapter04}
% ********************************************************************
% Backmatter
%*******************************************************
\appendix
%\renewcommand{\thechapter}{\alph{chapter}}
\cleardoublepage
%\part{Appendix}
%\include{Chapters/Chapter0A}
\printindex
%\cleardoublepage\include{FrontBackmatter/Bibliography}
%\cleardoublepage\include{FrontBackmatter/Declaration}
%\cleardoublepage\include{FrontBackmatter/Colophon}

\end{document}
